% Basic Category Theory
% Tom Leinster <Tom.Leinster@ed.ac.uk>
% 
% Copyright (c) Tom Leinster 2014-2016
% 
% Preface to the arXiv version
% 

\frontmatter

\setcounter{page}{3}


\chapter*{Preface to the arXiv version}


This book was first published by Cambridge University Press in 2014, and is
now being published on the arXiv by mutual agreement.  CUP has consistently
supported the mathematical community by allowing authors to make free
versions of their books available online.  Readers may, in turn, wish to
support CUP by buying the printed version, available at
\href{http://www.cambridge.org/9781107044241}{\url{http://www.cambridge.org/9781107044241}}.

This electronic version is not only free; it is also freely
\emph{editable}.  For instance, if you would like to teach a course using
this book but some of the examples are unsuitable for your class, you can
remove them or add your own.  Similarly, if there is notation that you
dislike, you can easily change it; or if you want to reformat the text for
reading on a particular device, that is easy too.

In legal terms, this text is released under the Creative Commons
Attribution-NonCommercial-ShareAlike 4.0 International licence (CC BY-NC-SA
4.0).  The licence terms are available at the Creative Commons website,
\href{https://creativecommons.org/licenses/by-nc-sa/4.0/}{\url{https://creativecommons.org/licenses/by-nc-sa/4.0}}.
Broadly speaking, the licence allows you to edit and redistribute this text
in any way you like, as long as you include an accurate statement about
authorship and copyright, do not use it for commercial purposes, and
distribute it under this same licence.

In technical terms, all you need to do in order to edit this book is to
download the source files from the \href{https://arxiv.org}{arXiv} and use
\LaTeX{} (or pdflatex) in the usual way.  

This version is identical to the printed version except for the correction
of a small number of minor errors.  Thanks to all those who pointed these
out, including Martin Brandenburg, Miguel Couto, Bradley Hicks, Thomas
Moeller, and Yaokun Wu.

\bigskip

\hfill\textit{Tom Leinster, December 2016}





\newpage

% Blank page

\thispagestyle{empty}

\  
